---
title: "rabies_model"
format: html
editor: visual
---

## Accelerating rabies elimination in Kenya

::: callout-note
We aimed to answer the following question:

-   What would be the impact of vaccination coverage, frequency and modification of dog demographics (culling / increased life span) in reaching the target of dog-mediated human rabies by 2030?
:::

### Overview

This model describes a deterministic compartmental model with the following compartments:

-   Susceptible - number of susceptibles dogs in the population

-   Exposed - number of exposed dogs in the population

-   Infectious - number of infectious dogs in the population

-   Vaccinated - number of successfully immunized dogs

The parameters for this model are:

-   lambda - the force of infection

-   carrying_capacity - the total dog capacity that is possible in this setting

-   beta - the transmission coefficient

-   progression_rate - the disease progression rate

-   rabies_mortality - the disease induced mortality rate

-   birth_rate - the per capita birth rate, which is estimated using the proportion of female dogs, female dogs of whelping age, annual rate of whelping and number of puppies per whelping

-   natural_mortality - the per capita backgroud mortality rate

-   importation_rate - the total importation rate (infectious dogs; assumed "0")

-   vaccination_coverage - the vaccination coverage

-   efficacy - the vaccine efficacy (proportion)

-   immunization_rate - the rate of immunized dogs per day; implemented as a wave during 100 days

-   loss_immunity - the rate at which dogs loose immunity afforded by prior vaccination

#### Vaccination in annual waves (60 days at the beginning of the year)

```{r, include=F}

library(deSolve)


seiv <- function(times,y,parms){
  with(c(as.list(y),parms),{
   
    N <- S + E + I + V 
    if(type==1){
  immunization_rate <- ifelse(times/365-floor(times/365)<0.272 & vaccination_coverage>=0.02,1/period_waves*vaccination_coverage,0.025/365)}
  else if(type==2){
    immunization_rate <-  ifelse(times/365-floor(times/365)<round(period_waves/365,3) &
                                  vaccination_coverage>=0.12,
                                1/365*vaccination_coverage,0.025/365)}
  else if(type==3){
    immunization_rate1 =  ifelse(times/365-floor(times/365)<round(period_waves/365,3) &
                                  vaccination_coverage>=0.02,
                                1/365*vaccination_coverage,0.025/365)
    immunization_rate2 =  ifelse(times/365-floor(times/365)<round(period_waves/365,3) &
                                  vaccination_coverage>=0.02,
                                1/period_waves*vaccination_coverage,0.025/(365))
    immunization <- data.frame(immunization_rate1, immunization_rate2)
    immunization$immunization_rate = immunization$immunization_rate1 + immunization$immunization_rate2
    immunization_rate <- immunization$immunization_rate}


    
      
    densityDep=TRUE # indicated that transmission is density-dependent
    lambda <-  ifelse(densityDep,beta*I/N0,beta*I/N)
    carrying_capacity=9000000 # carrying capacity
  
     dS <- birth_rate*(S+V) - natural_mortality* (1+N/carrying_capacity) * S - immunization_rate * S - lambda * S + loss_immunity * V  # We assume that mortality increases with increasing population size
    dE <- lambda * S - natural_mortality* (1+N/carrying_capacity) * E  - progression_rate * E
    dI <- importation_rate + progression_rate * E - natural_mortality* (1+N/carrying_capacity) * I - rabies_mortality * I
    dV <- immunization_rate * S - natural_mortality* (1+N/carrying_capacity) * V - loss_immunity * V
   
    return(list(c(dS, dE, dI, dV)))
  })
}



time <- 0  # Start time
N0 <- 8063804 # Start population 

popInit <- c(S = 7055804,
             E = 5000,
             I = 3000,
             V = 1000000)     #Initial conditions estimated from the data  

values <- c(beta = 0.4, 
            progression_rate = 1 / 22, 
            rabies_mortality = 1/ 3.1, 
            birth_rate = 0.003, 
            natural_mortality = 1/(1.85*365), # per capita backgroud mortality rate
            importation_rate = 0, # total importation rate (infectious dogs)
            vaccination_coverage = 0.70, # vac coverage at the end of 100 days campaign
            efficacy = 1, # vaccine efficacy (proportion)
            carrying_capacity=300000, # carrying capacity
            loss_immunity=1/(3*365),
            type=1, #1=annual waves, 2=annual coverage, 3=combined
            period_waves=100
            ) # loss of immunity from vaccination over time
   
```

```{r, echo=F}

times <- seq(0,(365*20),1)  # the model will run for a total of 20 years 

ts.seiv <- data.frame(lsoda(
  y = popInit,               # Initial conditions for population
  times = times,             # Timepoints for evaluation
  func = seiv,                   # Function to evaluate
  parms = values                # Vector of parameters
))



ts.seiv$year=ts.seiv$time/365
ts.seiv$calyear=ts.seiv$year+2023


ts.seiv$Ntot <- ts.seiv$S + ts.seiv$E + ts.seiv$I + ts.seiv$V
# convert data to long format for visualization
ts.seiv1 <- ts.seiv %>%
  pivot_longer(c(S,E,I,V), names_to="compartments", values_to="values") %>%
  mutate(compartments=recode(compartments, "S"="Susceptible", "E"="Exposed", "I"="Infectious", "V"="Vaccinated")) %>%
  group_by(calyear) %>%
  mutate(Ntot=sum(values)) %>%
  ungroup()


ggplot(ts.seiv1[ts.seiv1$compartments%in%"Infectious",], aes(x=calyear, y=values, color=compartments))+ geom_line(size=1)+theme_bw()+geom_vline(xintercept = 2030, linetype=2)+
  scale_fill_brewer(palette = "Set1")+labs(x="Year", y="Number of dogs", color="")


```

```{r, echo=F}

ggplot(ts.seiv1, aes(x=calyear, y=Ntot))+ geom_line( color="blue")+theme_bw()+geom_vline(xintercept = 2030, linetype=2)+labs(x="Year",y="Dog population")



```

```         
```

```{r, echo=F}

# This code creates a loop over different levels of vaccination coverage to estimate:
# (A) The number of infectious dogs over time
# (B) The year of elimination (defined as <1 infectious dog)

changevac = data.frame(Nr=rep(NA,length(times)))

for (i in c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0)) {

  values["vaccination_coverage"]=i
  
ts.seiv <- data.frame(lsoda(
  y = popInit,               # Initial conditions for population
  times = times,             # Timepoints for evaluation
  func = seiv,                   # Function to evaluate
  parms = values                # Vector of parameters
))
changevac2 <- cbind(changevac, i=ts.seiv$I)
changevac <- changevac2
}

ts.seiv$calyear=ts.seiv$time/365+2023

changevac <- cbind(ts.seiv$calyear,changevac)

names(changevac) = c("calyear", "none",0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0)

changevac1 <- changevac %>%
  pivot_longer(c(3:12), names_to="coverage", values_to="proportion") %>%
  mutate(coverage=as.numeric(coverage))

ggplot(changevac1, aes(x=calyear,y=proportion, color=factor(coverage)))+
  labs(x="Year", y="Number of infectious dogs")+
  geom_line()+theme_bw()+geom_vline(xintercept = 2030, linetype=2)+scale_fill_brewer(palette="Set1")




```

```{r, echo=F}


### Plot of the year of elimination ###
timeplot<- changevac1 %>%
  filter(proportion<1) %>%
  mutate(coverage=coverage*100)


ggplot(timeplot[timeplot$coverage>60,], aes(x=calyear, y=proportion, color=factor(coverage)))+
  geom_line()+
  theme_bw()

```

### Continuous vaccination

```{r, echo=F}

seiv <- function(times,y,parms){
  with(c(as.list(y),parms),{
   
    N <- S + E + I + V # calculation of the total population
    immunization_rate = ifelse(times/365-floor(times/365)<0.272 & vaccination_coverage>=0.025,1/365*vaccination_coverage,0.025/365) # This line allows for a "vaccination wave" lasting 100 days every beginning of a year.
    densityDep=TRUE # indicated that transmission is density-dependent
    lambda <-  ifelse(densityDep,beta*I/N0,beta*I/N)
    carrying_capacity=9000000 # carrying capacity
  
    dSdt <- birth_rate*(S+V) - natural_mortality* (1+N/carrying_capacity) * S - immunization_rate * S - lambda * S + loss_immunity * V  # We assume that mortality increases with increasing population size
    dEdt <- lambda * S - natural_mortality* (1+N/carrying_capacity) * E  - progression_rate * E
    dIdt <- importation_rate + progression_rate * E - natural_mortality* (1+N/carrying_capacity) * I - rabies_mortality * I
    dVdt <- immunization_rate * S - natural_mortality* (1+N/carrying_capacity) * V - loss_immunity * V
    list(c(dSdt, dEdt, dIdt, dVdt))
  })
}

time <- 0  # Start time
N0 <- 8063804 # Start population 

popInit <- c(S = 7055804,
             E = 5000,
             I = 3000,
             V = 1000000)     #Initial conditions estimated from the data  

values <- c(beta = 0.4, # the transmission coefficient
            progression_rate = 1 / 22, # progression rate
            rabies_mortality = 1 / 3.1, # disease induced mortality rate
            birth_rate = 0.004, # per capita birth rate:  0.428 (females) * 0.554 (welping age) *(1.26/365) welping rate *3.7 puppies per welping
            natural_mortality = 1/(1.85*365), # per capita backgroud mortality rate
            importation_rate = 0, # total importation rate (infectious dogs)
            vaccination_coverage = 0.37, # vac coverage at the end of 100 days campaign
            efficacy = 1, # vaccine efficacy (proportion)
            carrying_capacity=300000, # carrying capacity
            loss_immunity=1/(3*365)) # loss of immunity from vaccination over time
   
times <- seq(0,(365*20),1)  # the model will run for a total of 20 years 

ts.seiv <- data.frame(lsoda(
  y = popInit,               # Initial conditions for population
  times = times,             # Timepoints for evaluation
  func = seiv,                   # Function to evaluate
  parms = values                # Vector of parameters
))



ts.seiv$year=ts.seiv$time/365
ts.seiv$calyear=ts.seiv$year+2023


ts.seiv$Ntot <- ts.seiv$S + ts.seiv$E + ts.seiv$I + ts.seiv$V
# convert data to long format for visualization
ts.seiv1 <- ts.seiv %>%
  pivot_longer(c(S,E,I,V), names_to="compartments", values_to="values") %>%
  mutate(compartments=recode(compartments, "S"="Susceptible", "E"="Exposed", "I"="Infectious", "V"="Vaccinated")) %>%
  group_by(calyear) %>%
  mutate(Ntot=sum(values)) %>%
  ungroup()


# This code creates a loop over different levels of vaccination coverage to estimate:
# (A) The number of infectious dogs over time
# (B) The year of elimination (defined as <1 infectious dog)

changevac = data.frame(Nr=rep(NA,length(times)))

for (i in c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0)) {

  values["vaccination_coverage"]=i
  
ts.seiv <- data.frame(lsoda(
  y = popInit,               # Initial conditions for population
  times = times,             # Timepoints for evaluation
  func = seiv,                   # Function to evaluate
  parms = values                # Vector of parameters
))
changevac2 <- cbind(changevac, i=ts.seiv$I)
changevac <- changevac2
}

ts.seiv$calyear=ts.seiv$time/365+2023

changevac <- cbind(ts.seiv$calyear,changevac)

names(changevac) = c("calyear", "none",0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0)

changevac1 <- changevac %>%
  pivot_longer(c(3:12), names_to="coverage", values_to="proportion") %>%
  mutate(coverage=as.numeric(coverage))

ggplot(changevac1, aes(x=calyear,y=proportion, color=factor(coverage)))+
  labs(x="Year", y="Number of infectious dogs")+
  geom_line()+theme_bw()+geom_vline(xintercept = 2030, linetype=2)+scale_fill_brewer(palette="Set1")




```

#### Combined

```{r}

{r, echo=F}

seiv <- function(times,y,parms){
  with(c(as.list(y),parms),{
   
    N <- S + E + I + V # calculation of the total population
    immunization_rate = ifelse(times/365-floor(times/365)<0.272 & vaccination_coverage>=0.025,1/365*vaccination_coverage,0.025/365) # This line allows for a "vaccination wave" lasting 100 days every beginning of a year.
    densityDep=TRUE # indicated that transmission is density-dependent
    lambda <-  ifelse(densityDep,beta*I/N0,beta*I/N)
    carrying_capacity=9000000 # carrying capacity
  
    dSdt <- birth_rate*(S+V) - natural_mortality* (1+N/carrying_capacity) * S - immunization_rate * S - lambda * S + loss_immunity * V  # We assume that mortality increases with increasing population size
    dEdt <- lambda * S - natural_mortality* (1+N/carrying_capacity) * E  - progression_rate * E
    dIdt <- importation_rate + progression_rate * E - natural_mortality* (1+N/carrying_capacity) * I - rabies_mortality * I
    dVdt <- immunization_rate * S - natural_mortality* (1+N/carrying_capacity) * V - loss_immunity * V
    list(c(dSdt, dEdt, dIdt, dVdt))
  })
}

time <- 0  # Start time
N0 <- 8063804 # Start population 

popInit <- c(S = 7055804,
             E = 5000,
             I = 3000,
             V = 1000000)     #Initial conditions estimated from the data  

values <- c(beta = 0.4, # the transmission coefficient
            progression_rate = 1 / 22, # progression rate
            rabies_mortality = 1 / 3.1, # disease induced mortality rate
            birth_rate = 0.004, # per capita birth rate:  0.428 (females) * 0.554 (welping age) *(1.26/365) welping rate *3.7 puppies per welping
            natural_mortality = 1/(1.85*365), # per capita backgroud mortality rate
            importation_rate = 0, # total importation rate (infectious dogs)
            vaccination_coverage = 0.37, # vac coverage at the end of 100 days campaign
            efficacy = 1, # vaccine efficacy (proportion)
            carrying_capacity=300000, # carrying capacity
            loss_immunity=1/(3*365)) # loss of immunity from vaccination over time
   
times <- seq(0,(365*20),1)  # the model will run for a total of 20 years 

ts.seiv <- data.frame(lsoda(
  y = popInit,               # Initial conditions for population
  times = times,             # Timepoints for evaluation
  func = seiv,                   # Function to evaluate
  parms = values                # Vector of parameters
))



ts.seiv$year=ts.seiv$time/365
ts.seiv$calyear=ts.seiv$year+2023


ts.seiv$Ntot <- ts.seiv$S + ts.seiv$E + ts.seiv$I + ts.seiv$V
# convert data to long format for visualization
ts.seiv1 <- ts.seiv %>%
  pivot_longer(c(S,E,I,V), names_to="compartments", values_to="values") %>%
  mutate(compartments=recode(compartments, "S"="Susceptible", "E"="Exposed", "I"="Infectious", "V"="Vaccinated")) %>%
  group_by(calyear) %>%
  mutate(Ntot=sum(values)) %>%
  ungroup()


# This code creates a loop over different levels of vaccination coverage to estimate:
# (A) The number of infectious dogs over time
# (B) The year of elimination (defined as <1 infectious dog)

changevac = data.frame(Nr=rep(NA,length(times)))

for (i in c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0)) {

  values["vaccination_coverage"]=i
  
ts.seiv <- data.frame(lsoda(
  y = popInit,               # Initial conditions for population
  times = times,             # Timepoints for evaluation
  func = seiv,                   # Function to evaluate
  parms = values                # Vector of parameters
))
changevac2 <- cbind(changevac, i=ts.seiv$I)
changevac <- changevac2
}

ts.seiv$calyear=ts.seiv$time/365+2023

changevac <- cbind(ts.seiv$calyear,changevac)

names(changevac) = c("calyear", "none",0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0)

changevac1 <- changevac %>%
  pivot_longer(c(3:12), names_to="coverage", values_to="proportion") %>%
  mutate(coverage=as.numeric(coverage))

ggplot(changevac1, aes(x=calyear,y=proportion, color=factor(coverage)))+
  labs(x="Year", y="Number of infectious dogs")+
  geom_line()+theme_bw()+geom_vline(xintercept = 2030, linetype=2)+scale_fill_brewer(palette="Set1")



county_shapefile1 <- county_shapefile %>%
  mutate()

```
